---
title: Growth_and_phenotypic_assays
format: html
editor: visual
bibliography:
  - grateful-refs.bib
---

Loading required packages

```{r}
knitr::opts_chunk$set(echo = TRUE)
library("ggplot2")
library("Rmisc")
library("rstatix")
library("ggpubr")
library("dplyr")
library("grateful")
```

## Growth assays

### Faceted growth profiles of several genes

\[\[Supplementary Figure 5\]\]

```{r}
# to plot parasitaemia measured by flow cytometry

growthassay <-read.csv(file="./rawdata/growthassays_v200124.csv",header = T)
summ.growthassay <- summarySE(growthassay, measurevar="Parasitaemia", groupvars=c("Line","Treatment","Erythrocytic.cycle.number"), na.rm=TRUE)
pd<-position_dodge(0.1)
colorPalette<-c("gray20","red3")
pgrowth<-ggplot(summ.growthassay, aes(x=Erythrocytic.cycle.number, y=Parasitaemia, group=interaction(Line, Treatment))) +
geom_ribbon(aes(ymin=Parasitaemia-se, ymax=Parasitaemia+se, fill=Treatment),alpha=0.3,position=pd) +
geom_line(position=pd,alpha=0.75,aes(color=Treatment)) +
scale_fill_manual(values=colorPalette) +
scale_color_manual(values=colorPalette) +
theme(axis.title.x = element_text(size=10,color="gray20"),axis.title.y = element_text(size=10,color="gray20"),
  legend.title=element_blank(),
  legend.background=element_blank(), legend.position=c(0.95,0.1),
  legend.key = element_blank(), legend.key.width = unit(0.5,"cm"),
  axis.text=element_text(size=10,color="gray20"),
  axis.line = element_line(colour = "gray20"),
  axis.ticks.y = element_blank(),
  axis.ticks.x = element_line(colour = "gray20"),
  #strip.background =element_rect(fill="gray50"),
  strip.background =element_blank(),
  strip.text = element_text(colour ="gray20",hjust=0),
  panel.grid=element_blank(), 
  panel.background=element_blank()) +
xlab("Erythrocytic cycle number") +
ylab("Parasitaemia (%)") +
scale_y_log10(breaks = c(0.01,0.1,1,10), labels = c(0.01,0.1,1,10)) + 
annotation_logticks(sides = "l",color="gray50",outside=T,size=0.25,
                      long = unit(0.15, "cm"),mid = unit(0.1, "cm"),short = unit(0.05, "cm")) + # to get logticks
coord_cartesian(ylim=c(1e-2,NA),clip = "off") + # to get logticks on the outside
facet_wrap(~factor(Line, c("rom1","rom3","rom4","rom8","rom9","rom10","piplc","sub1","p230p")),nrow=4) # to facet by each gene
pgrowth
ggsave("./outplots/pgrowth.pdf",width = 5,height = 5)
write.csv(summ.growthassay,file="./outdata/summ.growthassay.csv")
```

### For sub1 \[Figure 2\]

```{r}
# to plot parasitaemia measured by flow cytometry

pd<-position_dodge(0.1)
colorPalette<-c("gray20","red3")
pgrowth_sub1<-ggplot(summ.growthassay %>% filter(Line == "sub1"), aes(x=Erythrocytic.cycle.number, y=Parasitaemia, group=interaction(Line, Treatment))) +
geom_ribbon(aes(ymin=Parasitaemia-se, ymax=Parasitaemia+se, fill=Treatment),alpha=0.3,position=pd) +
geom_line(position=pd,alpha=0.75,aes(color=Treatment)) +
scale_fill_manual(values=colorPalette) +
scale_color_manual(values=colorPalette) +
theme(axis.title.x = element_text(size=10,color="gray20"),axis.title.y = element_text(size=10,color="gray20"),
legend.title=element_blank(),
legend.background=element_blank(), legend.position=c(0.95,0.1),
legend.key = element_blank(), legend.key.width = unit(0.5,"cm"),
axis.text=element_text(size=10,color="gray20"),
axis.line = element_line(colour = "gray20"),
axis.ticks.y = element_blank(),
axis.ticks.x = element_line(colour = "gray20"),
#strip.background =element_rect(fill="gray50"),
strip.background =element_blank(),
strip.text = element_text(colour ="gray20",hjust=0),
panel.grid=element_blank(), 
panel.background=element_blank()) +
xlab("Erythrocytic cycle number") +
ylab("Parasitaemia (%)") +
scale_y_log10(breaks = c(0.01,0.1,1,10), labels = c(0.01,0.1,1,10)) + 
  annotation_logticks(sides = "l",color="gray50",outside=T,size=0.25,
                      long = unit(0.15, "cm"),mid = unit(0.1, "cm"),short = unit(0.05, "cm")) + # to get logticks
  coord_cartesian(ylim=c(1e-2,NA),clip = "off") #+ # to get logticks on the outside
#facet_wrap(~factor(Line, c("rom1","rom3","rom4","rom8","rom9","rom10","piplc","sub1","p230p")),nrow=4)
pgrowth_sub1
ggsave("./outplots/pgrowth_sub1.pdf",width = 5,height = 5)
```

## Number of merozoites in schizonts \[Figure 3\]

Counted from IFA images of RAP and mock treated *rom8-shiftiko* line

```{r}
rom8.merozoites<-read.csv("./rawdata/rom8_merozoites_v200124.csv")

pmerozoites_rom8<- ggplot(rom8.merozoites,
       aes(x=factor(Treatment),y=MerozoiteNo)) +
  geom_dotplot(aes(fill = Treatment),   # Use fill = Species here not in ggplot()
                 binaxis = "y",         # which axis to bin along
                 binwidth = 0.5,        # Minimal difference considered different
                 stackdir = "center",    # Centered
                 dotsize = 2,
                 color=NA
               ) +
stat_summary(fun = median, fun.min = median, fun.max = median,
                 geom = "crossbar", width = 0.25, colour = "gray20") +
  theme_bw() +
  scale_fill_manual(values = c("#BBBBBB","#CC3311")) +
  theme(axis.title = element_text(size=10,color="gray20"),
        axis.text=element_text(color="gray20"),
        axis.line = element_line(colour = "gray20"),
        panel.grid=element_blank(), 
        panel.background=element_blank(), 
        legend.position = "none") +
  xlab("Treatment") +
  ylab("Merozoite Number")

pwc <- rom8.merozoites %>%
  t_test(MerozoiteNo ~Treatment, p.adjust.method = "bonferroni") %>%
  add_xy_position(x = "Treatment", fun = "mean_sd", dodge = 0.8)

pmerozoites_rom8<- pmerozoites_rom8 + stat_pvalue_manual(
    pwc, label = "p", tip.length = 0.01,
    bracket.nudge.y = 5
    ) +
  scale_y_continuous(expand = expansion(mult = c(0.1, 0.1)))

#plotly does not work with geom_dotplot
pmerozoites_rom8
ggsave("./outplots/pmerozoites_rom8.pdf",width= 4, height= 4)
```

## Invasion assay \[Figure 3\]

4 hour invasion assays of RAP and mock treated *rom4-shiftiko* line under static and shaking conditions

```{r}
rom4.invasion <- read.csv("./rawdata/rom4_invasionassay_v200124.csv")

rom4.invasion.FC<-spread(rom4.invasion,Timepoint,Parasitaemia,fill=0) %>% transmute(Line,Treatment,Condition,Replicate,FoldChange=AFTER/BFOR)

pinvasion_rom4<-ggplot(rom4.invasion.FC,
       aes(x=factor(Treatment),y=FoldChange)) +
  geom_dotplot(aes(fill = Treatment),   # Use fill = Species here not in ggplot()
                 binaxis = "y",         # which axis to bin along
                 binwidth = 0.5,        # Minimal difference considered different
                 stackdir = "center",    # Centered
                 dotsize = 1,
                 color=NA
               ) +
stat_summary(fun = median, 
             fun.min = median, 
             fun.max = median,
             geom = "crossbar", 
             width = 0.25,
             colour = "gray20") +
  facet_grid(.~Condition,labeller=labeller(Condition=c(SHAK="SHAKING",STAT="STATIC"))) +
  theme_bw() +
  theme(axis.title = element_text(size=10,color="gray20"),
        axis.text=element_text(color="gray20"),
        axis.line = element_line(colour = "gray20"),
        panel.grid=element_blank(), 
        panel.background=element_blank(), 
        legend.position = "none") +
  scale_fill_manual(values = c("#BBBBBB","#CC3311")) +
  xlab("Treatment") + 
  ylab("Fold Change in Parasitaemia")

pwc <- rom4.invasion.FC %>% group_by(Condition) %>%
  t_test(FoldChange ~Treatment, p.adjust.method = "bonferroni") %>%
  add_xy_position(x = "Treatment", fun = "mean_sd", dodge = 0.8)

pinvasion_rom4 <- pinvasion_rom4 + stat_pvalue_manual(
    pwc, label = "p", tip.length = 0.01,
    bracket.nudge.y = 1
    ) +
  scale_y_continuous(expand = expansion(mult = c(0.1, 0.1)))
#plotly does not work with geom_dotplot
pinvasion_rom4
ggsave("./outplots/pinvasion_rom4.pdf",width= 4, height= 4)
```

## Citations

```{r}
cite_packages(output="paragraph",out.dir = getwd())
```
